\documentclass{article}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{geometry}
\geometry{margin=1in}

\title{Explanation of the Baboons Crossing Problem Solution}
\author{}
\date{}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{red},
  showstringspaces=false,
  frame=single,
}

\begin{document}

\maketitle

\section*{Problem Overview}

The problem describes a synchronization challenge where baboons attempt to cross a canyon using a single rope. The constraints are:

\begin{itemize}
    \item Baboons crossing in opposite directions must not meet on the rope.
    \item The rope supports at most 5 baboons at a time.
    \item No starvation: baboons waiting to cross in one direction should eventually get a chance even if there is a continuous stream in the opposite direction.
\end{itemize}

\section*{Code Explanation}

\subsection*{Global Declarations}

\begin{lstlisting}[language=C]
#define MAX_BABOONS 100
#define MAX_BABOONS_ON_ROPE 5

sem_t mutex;
sem_t leftQueue, rightQueue;

int count = 0;
int direction = 0;
int leftWaiting = 0;
int rightWaiting = 0;

char rope[MAX_BABOONS_ON_ROPE];
\end{lstlisting}

\textbf{Explanation:}
\begin{itemize}
    \item \texttt{MAX\_BABOONS} and \texttt{MAX\_BABOONS\_ON\_ROPE} define the total number of baboons and the rope capacity.
    \item \texttt{mutex}: a binary semaphore for mutual exclusion.
    \item \texttt{leftQueue}, \texttt{rightQueue}: semaphores for baboons waiting on each side.
    \item \texttt{count}: tracks the number of baboons currently on the rope.
    \item \texttt{direction}: tracks the current direction of crossing (\texttt{0} = none, \texttt{-1} = left-to-right, \texttt{1} = right-to-left).
    \item \texttt{leftWaiting}, \texttt{rightWaiting}: count baboons waiting to cross from each side.
    \item \texttt{rope[]}: visual representation of baboons on the rope.
\end{itemize}

\subsection*{Baboons Crossing Functions}

Two thread functions represent baboons crossing from each side.

\subsubsection*{Left to Right}

\begin{lstlisting}[language=C]
void* cross_left_to_right(void* arg) {
    sem_wait(&mutex);
    leftWaiting++;
    while (direction == 1 || count == MAX_BABOONS_ON_ROPE) {
        sem_post(&mutex);
        sem_wait(&leftQueue);
        sem_wait(&mutex);
    }
    ...
}
\end{lstlisting}

\textbf{Explanation:}
\begin{itemize}
    \item Baboons wait if direction is the opposite or the rope is full.
    \item Once safe, direction is set and \texttt{count} is incremented.
\end{itemize}

After crossing:

\begin{lstlisting}[language=C]
    count--;
    rope[count] = ' ';
    if (count == 0) {
        direction = 0;
        if (rightWaiting > 0) {
            for (int i = 0; i < rightWaiting && i < MAX_BABOONS_ON_ROPE; i++) {
                sem_post(&rightQueue);
            }
        } ...
    }
\end{lstlisting}

\textbf{Explanation:}
\begin{itemize}
    \item When no baboons are on the rope, the direction is reset.
    \item Baboons waiting on the opposite side are released, ensuring fairness (prevents starvation).
\end{itemize}

\subsubsection*{Right to Left}

\begin{lstlisting}[language=C]
void* cross_right_to_left(void* arg) {
    sem_wait(&mutex);
    rightWaiting++;
    while (direction == -1 || count == MAX_BABOONS_ON_ROPE) {
        sem_post(&mutex);
        sem_wait(&rightQueue);
        sem_wait(&mutex);
    }
    ...
}
\end{lstlisting}

\textbf{Explanation:}
\begin{itemize}
    \item Similar to left-to-right but for the opposite direction.
    \item Ensures only one direction uses the rope at a time and max 5 baboons.
\end{itemize}

\subsection*{Main Function}

\begin{lstlisting}[language=C]
int main() {
    sem_init(&mutex, 0, 1);
    sem_init(&leftQueue, 0, 0);
    sem_init(&rightQueue, 0, 0);

    for (int i = 0; i < MAX_BABOONS; i++) {
        if (rand() % 2 == 0)
            pthread_create(&baboons[i], NULL, cross_left_to_right, NULL);
        else
            pthread_create(&baboons[i], NULL, cross_right_to_left, NULL);
    }
    ...
}
\end{lstlisting}

\textbf{Explanation:}
\begin{itemize}
    \item Semaphores are initialized.
    \item Randomly creates threads representing baboons from both sides.
    \item Waits for all threads to finish and cleans up.
\end{itemize}

\section*{Mapping to Problem Constraints}

\begin{itemize}
    \item \textbf{Safety (no fights):} Direction is enforced via the \texttt{direction} variable. Baboons wait if the rope is used in the opposite direction.
    \item \textbf{Capacity (no breakage):} The \texttt{count} variable ensures no more than 5 baboons are on the rope.
    \item \textbf{No Starvation:} When rope is empty, baboons on the opposite side are signaled. If none, same-side baboons are allowed.
\end{itemize}

\end{document}
